# Assignment 1: Manta Disk Length

### Question 1(a): Why is a linear model more appropriate than other types of GLMs we’ve met in this course so far?

Variables which measure size such as width, length and height are typically normally
distributed when either any of them are the response variable and the rest are explanatory variables therefore it appropriate to assume that the response distribution for the disc width as explained by the explanatory variables: disc length, cranial width and sex is normally distributed.

### Question 1(b): Create plots to investigate relationships between the variables in this data set. Briefly comment on the plots (2–3 sentences).

```{r}
setwd("/Users/tarineccleston/Documents/Masters/STATS 762/regression-for-DS/manta-disc-width")
library(tidyverse)
library(ggplot2)

manta_df = read.csv("data/manta-measurements.csv")

# convert F to 0, and M to 1 for the normal regression model
manta_df$sex = gsub('F', 0, manta_df$sex)
manta_df$sex = gsub('M', 1, manta_df$sex)
manta_df$sex = as.numeric(manta_df$sex)

# fitting data
manta_fit = lm(DW ~ DL + CW + sex, data = manta_df)
summary(manta_fit)

# plotting data, best to use pairs plot as this shows the relationships between each
# variable
pairs(manta_df)
```

- Clear linear relationship between disc length and disc width, disc length and 
cranial width, and disc width and cranial width
- Greater variance of size for female manta rays compared to males if you compare 
the spread in the CW vs sex plot for example
- Male manta rays appear to be smaller than female manta rays

### Question 1(c): Create the model’s design matrix. Only print out the first six rows.

```{r}
# append a column of 1s at the beginning for the intercept. remove the response variable (disc width)
# make sure X and Y are in matrix form for the calculation
X = as.matrix(cbind(1, manta_df[,-2]))
head(X)
```

### Question 1(d): Calculate βb, the vector of estimated coefficients.

```{r}
# Calculate βb by using: betahat = (Xt*X)^-1 *Xt*Y
# Where X is the design matrix, and Y is the response variable matrix

Y = as.matrix(manta_df[,2])
# calculation
beta_hat = solve(t(X) %*% X) %*% t(X) %*% Y
colnames(beta_hat) = c("beta_hat")
beta_hat
```

### Question 1(e): Calculate μb, the vector of estimated expected values (or “fitted values”). Only print out the first six values.

```{r}
# calculate μb using: mew = X*betahat

mew = X %*% beta_hat
colnames(mew) = c("mew")
head(mew)
```

### Question 1(f): Calculate σb, the estimated error variance.

```{r}
# calculate σb using: σb^2 = RSS / (n - k - 1)
# where RSS = THETAsum (yi - f(xi))^2

# calculate RSS first
RSS = 0
Yhat = X %*% beta_hat
for (i in 1:nrow(X)) {
  diff_i = Y[i] - Yhat[i]
  RSS = RSS + (diff_i)^2
}

var_hat_sq = RSS/(nrow(X) - ncol(X))
var_hat_sq
```


### Question 1(g): Calculate standard errors for the estimated coefficients.

```{r}
cov = var_hat_sq*solve(t(X)%*%X)
std_err = sqrt(diag(cov))
std_err
```

### Question 1(h): Calculate t-test statistics and p-values to test individual null hypotheses that the coefficients are equal to zero.

```{r}
t_stat = beta_hat/std_err
colnames(t_stat) = c("t-stat")
t_stat
p_value = 2*(1-pt(t_stat, 20))
colnames(p_value) = c("p-value")
p_value
```

### Question 1(i): Calculate confidence intervals for the coefficients.

```{r}
beta_CI = cbind(std_err - qt(0.975, df = 20)*std_err, beta_hat + qt(0.975, df = 20)*std_err)
colnames(beta_CI) = c("2.5%", "97.5%")
beta_CI
```

### Question 1(j): Calculate point predictions (equivalent to estimated expectations) for Morag’s and Evelyn’s disc widths.

```{r}
# Morag: a female manta ray with a disc length of 1.3 m and a cranial width of 0.75 m.
# Evelyn: a female manta ray with a disc length of 1.625 m and a cranial width of 0.9375 m.

# build new design matrix based on the given data
X_pred = cbind(c(1, 1), c(1.3, 1.625), c(0.75, 0.9375), c(0, 0))
rownames(X_pred) =  c("Evelyn", "Morag")
colnames(X_pred) = c("1", "DL", "CW", "sex")
X_pred

# getting prediction
Y_pred = X_pred %*% beta_hat
rownames(Y_pred) =  c("Evelyn", "Morag")
colnames(Y_pred) = c("DW")
Y_pred
```

### Question 1(k): Calculate confidence intervals for Morag’s and Evelyn’s expected disc widths.

Calculate prediction variance and standard errors from new design matrix

```{r}
pred_var = as.matrix(diag(X_pred %*% cov %*% t(X_pred)))
pred_SE = sqrt(pred_var)

pred_CI = cbind(Y_pred - qt(0.975, df = 20)*pred_SE, Y_pred + qt(0.975, df = 20)*pred_SE)
colnames(pred_CI) = c("2.5%", "97.5%")
pred_CI
```

### Question 1(l) Calculate prediction intervals for Morag’s and Evelyn’s observed disc widths.

```{r}
# a prediction interval captures the uncertainty around a single value. 
# a confidence interval captures the uncertainty around the mean predicted values

# combine variances and calculate combined standard error
comb_var = pred_var + var_hat_sq
comb_SE = sqrt(comb_var)

pred_PI = cbind(Y_pred - qt(0.975, df = 20)*comb_SE, Y_pred + qt(0.975, df = 20)*comb_SE)
colnames(pred_PI) = c("2.5%", "97.5%")
pred_PI
```

### Question 1(m) Bonus question! Write your own R function that accepts the arguments

```{r}
# Function Description: Basically does the same as the R summary() function
# input: 
#        Y, a vector of observations
#        X, a design matrix
# output:
#        estimated coefficients, 
#        standard errors 
#        t-test statistics
#        p-values

summary_custom <- function(X, Y) {
  # estimate coefficients
  # ______________________
  beta_hat = solve(t(X) %*% X) %*% t(X) %*% Y
  colnames(t_stat) = c("estimates")
  
  # standard error
  # ______________________
  # calculate RSS first
  RSS = 0
  Yhat = X %*% beta_hat
  for (i in 1:nrow(X)) {
    diff_i = Y[i] - Yhat[i]
    RSS = RSS + (diff_i)^2
  }
  
  var_hat_sq = RSS/(nrow(X) - ncol(X))
  cov = var_hat_sq*solve(t(X)%*%X)
  std_err = sqrt(diag(cov))
  colnames(t_stat) = c("stderr")
  
  # t-test statistics
  # ______________________
  t_stat = beta_hat/std_err
  colnames(t_stat) = c("t-stat")
  
  # p-value
  # ______________________  
  p_value = 2*(1-pt(t_stat, 20))
  colnames(p_value) = c("p-value")
  
  return(beta_hat, std_err, t_stat, p_value)
}
```

### Question 2(a) Under the model from Question 1, what is the estimated ratio of Evelyn’s expected disc width to Morag’s expected disc width?

```{r}
DW_ratio =  round((Y_pred[2] / Y_pred[1] - 1) * 100, digits = 3)
DW_ratio
```
