---
title: "Untitled"
output: html_document
date: "2023-06-02"
---

## STATS 762: Gaussian Process

```{r setup, include=FALSE}
setwd("/Users/tarineccleston/Documents/Masters/STATS 762/regression-for-DS")
library(tidyverse)
library(kernlab)
library(splines)

car_df = read_csv("data/car.csv")
glimpse(car_df)
```

### Task 1: Fit a Gaussian process regression.

```{r}
# remove missing values
car_df <- na.omit(car_df)
gauss_car_fit = gausspr(car_df[,-1], as.matrix(car_df$Price), variance.model = TRUE)
gauss_car_fit
```

### Task 2: Fit a B-spline regression with degree of 3. We assume the same number of knots for all explanatory variables and only consider 1-10 knots for simplicity.

```{r}
set.seed(103)

#LOOCV error -vs- number of knots
n.knots=c(0:10)

price.cv.bs = rep(0,length(n.knots))

for(j in 1:length(n.knots)){ for(l in 1:length(car_df)){
  #predict prices
  price.bs = predict(lm(Price ~ bs(HP, df = n.knots[j]+3, intercept=FALSE, Boundary.knots = range(HP)) + 
                          bs(Wheel.base, df = n.knots[j]+3, intercept=FALSE, Boundary.knots= range(Wheel.base)) +
                          bs(Disp, df = n.knots[j]+3, intercept=FALSE, Boundary.knots = range(Disp)), data = car_df[-l,]), newdata= car_df[l,])
  
  #cumulative sum of error
  price.cv.bs[j]=price.cv.bs[j]+(car_df$Price[l]-price.bs)^2
}}

# average cv error
price.cv.bs = price.cv.bs/length(car_df$Price)

plot(n.knots, log(price.cv.bs), ylab='log(cv error)',xlab='Number of knots', 'o')
```

3 knots seem best

### Task 3: We have fitted a regression tree for this data in Week 10. Compare the performance of nonparametric model in (a) , spline regression in (b) and your best regression tree in Week 10.  Which one is better? What can we say about the association between price and car specification?