---
title: "Untitled"
output: html_document
date: "2023-05-22"
---

## STATS 762: Assignment 4
### Author: Tarin Eccleston

#### Cardiovascular diseases (CVDs) are one of major causes of death glob-
ally. About 17.9 million lives die each year and this is about 31% of all
deaths worldwide. Heart failure is a common event caused by CVDs and
the dataset train.csv contains 10 features that can be used to predict
mortality by heart failure.

```{r}
setwd("/Users/tarineccleston/Documents/Masters/STATS 762/regression-for-DS/disease-and-weather/")
library(rpart)
library(randomForest)
library(tidyverse)
```

#### 1a) Which explanatory variables are nominal? Which explanatory vari-
ables are numeric? Identify them and modify data to read explana-
tory variables correctly. 

```{r}
cvd_df = read_csv("data/train.csv")
glimpse(cvd_df)
```

Nominal: anaemia, diabetes, high_blood_pressure, sex, smoking, death
Numeric: age, creatinine_phosphokinase, ejection_fraction, platelets, serum_sodium

```{r}
# force to factor
cvd_df$anaemia = as.factor(cvd_df$anaemia)
cvd_df$diabetes = as.factor(cvd_df$diabetes)
cvd_df$high_blood_pressure = as.factor(cvd_df$high_blood_pressure)
cvd_df$sex = as.factor(cvd_df$sex)
cvd_df$smoking = as.factor(cvd_df$smoking)
cvd_df$death = as.factor(cvd_df$death)
```

#### 1b) Fit an appropriate regression model using all explanatory variables.
Write the reason for your model choice. 

```{r}

```


