---
title: "STATS 762 - Slides demo"
output: html_document
---


## Quantile regression

First I loaded packages.
```{r}
library(quantreg)
library(pscl)
```

### Engel food data

Find quantile regressions for q=(0.05, ..., 0.95).
```{r}
data(engel)
T=c(0.05,0.1,0.25,0.5,0.75,0.9,0.95)
plot(foodexp ~ income, data = engel, main = "foodexp ~ income"
                ,ylab="Food Expenditure",xlab='Household Income')
abline(lm(foodexp ~ income, data = engel), col = 1); 

for(t in 1:length(T)){
  abline(rq(foodexp ~ income, data = engel,tau=T[t]), col = t+1)
}
legend("topright", legend = c("lm", paste("rq,",as.character(T))), col = c(1:(length(T)+1)),lty = 1)

```

Create table of intercept and coefficient estimates. (Slide 15)
```{r}
T=c(0.05,0.1,0.25,0.5,0.75,0.9,0.95)
S0=S1=matrix(,length(T),3)
for(t in 1:length(T)){
  S0[t,]=summary(rq(foodexp ~ income, data = engel,tau=T[t]))$coefficients[1,]
  S1[t,]=summary(rq(foodexp ~ income, data = engel,tau=T[t]))$coefficients[2,]
}

par(mfrow = c(1, 2)) 
plot(T,S0[,1],type='l',ylim=c(min(S0),max(S0)),xlab='quantiles',ylab='intercept',
     cex.axis=1.5,cex=1.5,cex.lab=1.5); 
lines(T,S0[,2],col='blue'); lines(T,S0[,3],col='blue')
plot(T,S1[,1],type='l',ylim=c(min(S1),max(S1)),xlab='quantiles',ylab='coefficient',
     cex.axis=1.5,cex=1.5,cex.lab=1.5); 
lines(T,S1[,2],col='blue'); lines(T,S1[,3],col='blue')

```

```{r}
rqfit <- rq(foodexp ~ income, data = engel,tau=0.9)
predict(rqfit,data.frame(income=1520),interval='confidence')

rqfit$coefficients[1]+rqfit$coefficients[2]*1520

```

### Weight-height data

```{r}
hw.df <- read.csv('weightheight.csv',header=TRUE)
plot(Weight ~ Height, data = hw.df, pch = substr(Gender, 1, 1), cex = 0.7,col = ifelse(Gender == "Male", "blue", "red"))
```

```{r}
hw.fit1 <- lm(Weight ~ Height*Gender, data = hw.df)
AIC(hw.fit1)
hw.fit2 <- lm(Weight ~ Height+Gender, data = hw.df)
AIC(hw.fit2)
hw.fit3 <- lm(Weight ~ Height, data = hw.df)
AIC(hw.fit3)
hw.fit4 <- lm(Weight ~ Gender, data = hw.df)
AIC(hw.fit4)
```
The 2nd model is chosen using AIC.

```{r}
summary(hw.fit2)
par(mfrow = c(2, 2)); plot(hw.fit2)
```

Average weights are predicted for male and female students.

```{r}
male <- data.frame(Height=seq(from=54,to=79,length.out=100),Gender='Male')
female <- data.frame(Height=seq(from=54,to=79,length.out=100),Gender='Female')
pred.male=predict(hw.fit2,male,interval = "confidence")
pred.female=predict(hw.fit2,female,interval = "confidence")

#changing the base line
hw.df$Gender <- factor(hw.df$Gender,levels=c("Male","Female"))
summary(lm(Weight ~ Height+Gender, data = hw.df))

plot(male$Height,pred.male[,1],type='l',ylim=c(80,250),ylab='Weight',xlab='Height');
lines(male$Height,pred.male[,2],lty=3);lines(male$Height,pred.male[,3],lty=3);
lines(female$Height,pred.female[,2],col=2);
lines(female$Height,pred.female[,2],lty=3,col=2);lines(female$Height,pred.female[,3],lty=3,col=2);
legend("topleft",c('Male','Female'),lty=c(1,1),col=c(1,2));
text(60,180,'Weight(Male) = -241.74 + 6.20*Height')
text(70,130,'Weight(Female) = -259.73 + 6.20*Height')
```

Quantiles of weights are predicted using quantile regressions.

```{r}
ql10 <- rq(Weight ~ Height+Gender, data = hw.df, tau=0.1)
summary(ql10)
ql90 <- rq(Weight ~ Height+Gender, data = hw.df, tau=0.9)
summary(ql90)

q10.pred.male=predict(ql10,male)
q10.pred.female=predict(ql10,female)
q90.pred.male=predict(ql90,male)
q90.pred.female=predict(ql90,female)

plot(male$Height,pred.male[,1],type='l',ylim=c(80,250),ylab='Weight',xlab='Height',main='Male students');
lines(male$Height,q10.pred.male,lty=3);lines(male$Height,q90.pred.male,lty=3,col=4);
text(70,130,'Weight(10%) = -239.55 + 5.98*Height')
text(60,180,'Weight(90%) = -215.93 + 6.04*Height')
legend("topleft",c('10%','90%'),lty=c(3,3),col=c(1,4));

plot(female$Height,pred.female[,1],type='l',ylim=c(80,250),ylab='Weight',xlab='Height',main='Female students');
lines(female$Height,q10.pred.female,lty=3);lines(female$Height,q90.pred.female,lty=3,col=4);
legend("topleft",c('10%','90%'),lty=c(3,3),col=c(1,4));
text(70,130,'Weight(10%) = -259.45 + 5.98*Height')
text(60,180,'Weight(90%) = -236.11 + 6.04*Height')

predict(ql10,data.frame(Height=c(70,70),Gender=c('Male','Female')),interval='confidence')

predict(ql90,data.frame(Height=c(70,70),Gender=c('Male','Female')),interval='confidence')

```

## Zero-inflated count data

Load the pscl-package and the fish caught data. 

```{r}
library(pscl)

fish <- read.csv('fish.csv',header=TRUE)
```

Fish count is modelled by the Poisson density.

```{r}
f0 <- c(0:150) #vector of count values
hist(fish$count,prob=TRUE,nclass=70,xlab='count'); lines(f0,dpois(f0,3.296))
```
### Zero-inflated Poisson regression 

Fish count is modelled by the zero-inflated Poisson model.

```{r}
#zero-inflated poisson dist
zinf1 <- zeroinfl(count~1,data=fish,dist='poisson')
summary(zinf1)
coef(zinf1)

#Back transformation for a zero-inflation probability 
p.zi = exp(coef(zinf1)[2])/(1+exp(coef(zinf1)[2])) 

hist(fish$count,prob=TRUE,nclass=70,xlab='count'); lines(f0,p.zi*(f0==0)+(1-p.zi)*dpois(f0,coef(zinf1)[1]),col=2)

#probability of zero catch 
p.zi+(1-p.zi)*dpois(0,exp(coef(zinf1)[1]))

```
Fish count for a number of adults and child is modelled using a zero-inflated Poisson regression. Particularly 

```{r}
zinf2 <- zeroinfl(count~persons+child,data=fish,dist='poisson')
```

Predict a number of fish caught for a group of 1 adult and 2 children.
```{r}
#zero-inflated poisson dist
exp(coef(zinf2)[4]+coef(zinf2)[5]*3+coef(zinf2)[6]*2)/(1+exp(coef(zinf2)[4]+coef(zinf2)[5]*3+coef(zinf2)[6]*2))

predict(zinf2,data.frame(persons=3,child=2),type='zero')

#expected non-zero value (mean for untruncated process)
predict(zinf2,data.frame(persons=3,child=2),type='count')

#expected value
predict(zinf2,data.frame(persons=3,child=2),type='response')

```

### Hurdle model

```{r}
zinf3 <- hurdle(count~persons+child,data=fish,dist='poisson',link='logit')
summary(zinf3)
```
Predict a number of fish caught for a group of 1 adult and 2 children.
```{r}
#mean parameter for untruncated process
predict(zinf3,data.frame(persons=3,child=2),type='count')

#expected value
predict(zinf3,data.frame(persons=3,child=2),type='response')

#ratio of non-zero term
predict(zinf3,data.frame(persons=3,child=2),type='zero')

```